---
title: "Introduction to Data Science in R"
author: "Mehman Ismayilli"
date: "2023-08-17"
output:
  pdf_document: default
  pdf: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R basics


```{r}
library(dslabs)
data("murders")
```

### Sorting
From page 61

`sort` function
```{r}
sort(murders$total)
```
`order` function
```{r}
index <- order(murders$total) ## generates index for murders
states <- murders$abb[index] ## finds abbreviations
```

`max` and `which.max`
```{r}
max(murders$total)
#find the index number
ind_max <- which.max(murders$total)
#find the particular state
murders$state[ind_max]
```
`rank`
```{r}
rank(murders$total)
```

### Vector arithmetics

California has the highest murder number. Is it so dangerous in California? To do so, let us find murder rate per 100 K in each state
```{r}
murder_rate <- murders$total/murders$population * 100000
```

Now find the state with the highest murder rate for 100 K

```{r}
murders$abb[order(murder_rate)]
```

You can see that the DC has the highest murder rate for 100 K.

### Indexing

Now indexing based on a criterion

```{r}
ind <- murder_rate < 0.71
murders$state[ind]

#change the condition 
ind <- murder_rate <= 0.71
murders$state[ind]

#count how many in there
sum(ind)
```

`which` function indexing

```{r}
ind <- which(murders$state=="California")
murder_rate[ind]
```

`match`
```{r}
ind <- match(c("New York", "Florida", "Texas"), murders$state)
murder_rate[ind]
```

`%in$`
To check whether a set contains certain elements
```{r}
c("Boston", "Dakota", "Washington") %in% murders$state
```

### Exercise 2.14
```{r}
low <- murder_rate < 1
ind <- which(low == TRUE)
murders$state [ind]
```


```{r}
ind <- which(low == TRUE & murders$region=="Northeast")
murders$state[ind]
```

```{r}
ind <- which(murder_rate < mean(murder_rate))
length(ind)
```

```{r}
ind <- match(c("AK", "MI", "IA"), murders$abb)
murders$state[ind]
```

```{r}
c("MA", "ME", "MI", "MO", "MU") %in% murders$abb
```

### Basic plots

```{r}
x <- murders$population / 10^6
y <- murders$total
plot(x, y)
```

For a quick plot that avoids accessing variables twice, we can use `with` function

```{r}
with(murders, plot(population, total))
```

histogram

```{r}
x <- with(murders, total/population * 100000)
hist(x)
```
```{r}
murders$state[which.max(x)]
```

`boxplot`
```{r}
boxplot(murder_rate~region, data = murders)
```

`image`
```{r}
x <- matrix(1:120, 12, 10)
image(x)
```

## Chapter 3 Programming basics

```{r}
a <- 0

if(a!=0){
  print(1/a)
} else{
  print("No reciprocal for 0.")
}
```

```{r}
ind <- which.min(murder_rate)

if (murder_rate[ind] <0.5){
  print(murders$state[ind])
} else {
  print("No state has murder rate that below")
}
```

A shorter version of `if` function, `ifelse`
```{r}
a <- 0
ifelse(a>0, 1/a, NA)
```

```{r}
a <- c(0, 1, 2, -4, 5)
result <- ifelse(a>0, 1/a, NA)
result
```

```{r}
data("na_example")
no_nas <- ifelse(is.na(na_example), 0, na_example)
sum(is.na(no_nas))
```

`any` and `all` functions
```{r}
z <- c(TRUE, TRUE, FALSE)
any(z)
all(z)
```

### Defining functions
```{r}
avg <- function(x){
  s <- sum(x)
  n <- length(x)
  s/n
}
```

```{r}
x <- 1:100
avg(x)
identical(mean(x), avg(x))
```
A bit complex function
```{r}
avg <- function(x, arithmetic=TRUE){
  n <- length(x)
  ifelse(arithmetic, sum(x)/n, prod(x)^(1/n))
}
avg(x, arithmetic = FALSE)
```

### For-loops
```{r}
compute_s_n <- function(n){
  x <- 1:n
  sum(x)
}
compute_s_n(100)
```

```{r}
for(i in 3){
  print(i)
}
```
```{r}
m <- 25
s_n <- vector(length = m) #create an empty vector
for(n in 1:m){
  s_n[n] <- compute_s_n(n)
}
```

```{r}
n <- 1:m
plot(n, s_n)
```

### Vectorisation and functionals

```{r}
x<- 1:10
sqrt(x)
y <- 1:10
x*y
```

The following piece of code does not run the function on each entry of `n`. 
```{r}
n <- 1:25
compute_s_n(n)
```

```{r}
x <- 1:10
sapply(x, sqrt)
```

fix the problem above
```{r}
n <- 1:25
s_n <- sapply(n, compute_s_n)
s_n
```

## The tidyverse
```{r}
library(tidyverse)
```
Create a new column by `mutate`
```{r}
murders <- mutate(murders, rate = total/population*100000)
```

Subsetting with `filter`
```{r}
filter(murders, rate <=0.81)
```

Selecting columns with `select` and create a new data tables

```{r}
new_table <- select(murders, state, region, rate)
filter(new_table, rate <= .71)
```

```{r}
murders <- mutate(murders, pop_in_millions = population/10^6)
```

```{r}
select(murders, state, population) %>%head()
```

```{r}
filter(murders, state == "New York")
```
Choose the data excluding "Florida"
```{r}
no_florida <- filter(murders, state != "Florida")
```

```{r}
filter(murders, population < 5000000 & region == "Northeast")
```

```{r}
filter(murders, state %in% c("New York", "Texas"))
```

### The pipe `%>%`
```{r}
murders %>% select(state, region, rate) %>% filter(rate <= .71)
```

```{r}
16 %>% sqrt() %>% log2()
```

### Summarising data

```{r}
data("heights")
```
Average heights and st dev of females
```{r}
s <- heights %>% 
  filter(sex == "Female") %>%
  summarise(average = mean(height), st_dev = sd(height))
s
```


```{r}
s$average
s$st_dev
```

```{r}
heights %>% filter(sex == "Female") %>%
  filter(sex == "Female") %>%
  summarise(median = median(height), 
            minimum = min(height), 
            maximum = max(height))
```

```{r}
heights %>% 
  filter(sex == "Female") %>%
  summarise(range = quantile(height, c(0, 0.5, 1)))
```

US murder rate
```{r}
us_murder_rate <- murders %>%
  summarise(rate = sum(total)/sum(population)*100000)
us_murder_rate
```
`pull`

```{r}
class(us_murder_rate)
```
`dplyr` returns outcome in a dataframe but we want to `pull` the output
```{r}
us_murder_rate %>% pull(rate)
```
which is equivalent of `us_murder_rate$rate`. To get the value immediated
```{r}
us_murder_rate <- murders %>%
  summarise(rate = sum(total)/sum(population)*100000) %>% 
  pull(rate)
us_murder_rate
```

### Group then summarise with `group_by`

Split data into *groups* and then compute summaries for each group.

```{r}
heights %>% group_by(sex) %>%
  summarise(average = mean(height), standard_deviation = sd(height))
```
```{r}
murders %>% group_by(region) %>%
  summarise(median_rate = median(rate))
```

### Sorting data frames
```{r}
murders %>% arrange(population) %>% head()
```

```{r}
murders %>% arrange(rate) %>% head()
```

`top_n` function
Top `n` ordering, e.g., top 5 highest murder rate
```{r}
murders %>% top_n(5, rate)
```

```{r}
murders %>% top_n(5, desc(rate))
```


```{r}
library(NHANES)
data(NHANES)
```

```{r}
ref <- NHANES %>% filter(AgeDecade == " 20-29") %>%
          summarise(average = mean(BPSysAve, na.rm=TRUE), st_dev = sd(BPSysAve, na.rm = TRUE)) 
ref
```
```{r}
ref_avg <- ref %>% pull(average)
ref_avg
```

```{r}
NHANES %>% filter(AgeDecade == " 20-29") %>% 
  summarise(max_pressure = max(BPSysAve, na.rm = TRUE), 
            min_pressure = min(BPSysAve, na.rm = TRUE)) %>%
  pull(max_pressure, min_pressure)
```

```{r}
NHANES %>% filter(Gender == "female") %>%
  group_by(AgeDecade) %>%
  summarise(average = mean(BPSysAve, na.rm=TRUE), st_dev = sd(BPSysAve, na.rm = TRUE))
```

```{r}
NHANES %>% filter(Gender == "male") %>%
  group_by(AgeDecade) %>%
  summarise(average = mean(BPSysAve, na.rm=TRUE), st_dev = sd(BPSysAve, na.rm = TRUE))
```

`do` function

```{r}
my_summary <- function(dat){
  x <- quantile(heights$height, c(0, 0.5, 1))
  tibble(min=x[1], median = x[2], max = x[3])
}
```

```{r}
heights %>% 
  group_by(sex) %>%
  my_summary
```
However, we want this outcome for each sex; therefore, we can use `do` function instead
```{r}
heights %>% 
  group_by(sex) %>%
  do(my_summary(.))
```
if we do not use dot then it would not return any outcome. 

### `purr` package
It provides much advanced version of `sapply` functions to receive each ourput withing specified classification. 
```{r}
library("purrr")
```

```{r}
n <- 25
s_n <- sapply(n, compute_s_n)
class(s_n)
```
`map()` works as `sapply` but always returns list
```{r}
s_n <- map(n, compute_s_n)
s_n
```
for a numeric vector 
```{r}
s_n <- map_dbl(n, compute_s_n)
s_n
```

### Tidyverse conditionals
`case_when`

```{r}
x <- c(-2, -1, 0, 1, 2)
case_when(x<0 ~ "Negative", x>0 ~ "Positive", TRUE ~ "Zero")
```

A more advanced but practical example showing how states are located at different regions
```{r}
murders %>% 
  mutate(group = case_when(
    abb %in% c("ME", "NH", "VT", "MA", "RI", "CT") ~ "New England",
    abb %in% c("WA", "OR", "CA") ~ "West Coast", 
    region == "South" ~ "South",
    TRUE ~ "Other")) %>%
  group_by(group) %>%
  summarise(rate = sum(total)/sum(population)*10^5)
```

### Exercise 4.15
part 1 
```{r}
is_tibble(murders)
class(murders)
```

part 2 
```{r}
murders_tibble <- as_tibble(murders)
```

part 3
```{r}
murders %>% 
  group_by(region)
```

part 4
```{r}
murders %>%
  pull(population) %>% log %>% mean %>% exp

```

## Importing data

### Paths and the working directory
```{r}
filename <- "murders.csv"
dir <- system.file("extdata", package = "dslabs")
fullpath <- file.path(dir, filename)
file.copy(fullpath, "murders.csv")
```

Working directory
```{r}
wd <- getwd()
```

## ggplot2
```{r}
library(ggplot2)
```

Now let us develop a graph based step-by-step by calling `ggplot` objects
```{r}
#first define the data
murders %>% ggplot() #alternatively, ggplot(data= murders)
```
No geometry is defined therefore it produces only a blank gray canvas. Now let us assign the plot to an object

```{r}
p <- murders %>% ggplot()
class(p)

p <- murders %>% ggplot() +
  geom_point(aes(x=population/10^6, y=total)) 

p
#now label the points
p+geom_text(aes(x=population/10^6, y=total, label= abb))
```

Final graph with all codes
```{r}
p <- murders %>% ggplot() +
  geom_point(aes(x=population/10^6, y=total), size = 3) +
  geom_text(aes(x=population/10^6, y=total, label=abb), nudge_x = 1.5)
p
```

Previously we defined the mapping `aes()` twice, locally. However, let us now define it _globally_.

```{r}
p <- murders %>% ggplot(aes(population/10^6, total, label = abb)) +
  geom_point(size=3) + 
  geom_text(nudge_x = 1.5)

p
```
_Scales_
```{r}
p <- murders %>% ggplot(aes(population/10^6, total, label = abb)) +
  geom_point(aes(col=region),size=3) + 
  geom_text(nudge_x =0.05)+
  scale_x_log10() + #scale_x_continuous(trans = "log10") +
  scale_y_log10()+  #scale_y_continuous(trans = "log10")
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010")

p
```

Annotation, shapes and adjustments
```{r}
r <- murders %>% 
  summarise(rate = sum(total)/sum(population)*10^6) %>%
  pull(rate)

#Now add the line into the graph
p+geom_abline(intercept = log10(r), lty=2, color = "darkgrey")

p <- p+scale_color_discrete(name="Region") #changes color legend title
```

### Add on packages for `ggplot`
```{r}
library(ggthemes)
library(ggrepel)
```

```{r}
p<- p+theme_economist()
p
```

#### Putting it all together

```{r}
r <- murders %>% 
  summarise(rate = sum(total)/sum(population)*10^6) %>%
  pull(rate)

p <- murders %>% ggplot(aes(population/10^6, total, label=abb))+
  geom_abline(intercept = log10(r), lty=2, color = "darkgrey") +
  geom_point(aes(col=region), size=3) +
  geom_text_repel() +
  scale_x_log10() +
  scale_y_log10() + 
  xlab("Populations in millions (log scale)") + 
  ylab("Total number of murders (log scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region") + 
  theme_economist()
p

```

`qplot` was decprecated in ggplots2 3.4.0; therefore, my suggestion is to use `plot` per se.

```{r}
plot(log10(murders$population), murders$total)
```

```{r}
library(gridExtra)
x <- log10(murders$population)
y <- murders$total
p1 <- qplot(x,y)
p2 <- qplot(x)
grid.arrange(p1, p2, ncol=2)
```

## Visualising data distributions

Frequency table
```{r}
heights %>% count(sex) %>% mutate(proportion = n/sum(n))
```

```{r}
index <- heights$sex == "Male"
x <- heights$height[index]

c(mean = mean(x), st_dev = sd(x))
```

Scaling 
```{r}
z <- scale(x)
mean(abs(z)<2)
```

Quantile-quantile plots
```{r}
pnorm(-1.96)
```

`qnorm`
```{r}
qnorm(0.975)
```

```{r}
qnorm(.975, mean = 5, sd=2)
```
```{r}
p <- seq(.05, .95, .05)
sample_quantiles <- quantile(x, p)
sample_quantiles
```

```{r}
theoretical_quantiles <- qnorm(p, mean(x), sd(x))
```

```{r}
qplot(x=theoretical_quantiles, y=sample_quantiles) + geom_abline()
```

```{r}
heights %>% filter(sex== "Male") %>% ggplot(aes(sample=scale(height))) +
                                              geom_qq() + geom_abline()
```

```{r}
murders %>% ggplot(aes(region))+geom_bar()
```


```{r}
tab <- murders %>% count(region) %>% mutate(proportion = n/sum(n))
tab
tab %>% ggplot(aes(region, proportion))+geom_bar(stat = "identity")
```

```{r}
heights %>% filter(sex == "Female") %>% ggplot(aes(height)) + 
  geom_histogram(binwidth = 1, fill="blue", col = "black")+
  xlab("Male heights in inches")+
  ggtitle("Histogram")
```

```{r}
heights %>% filter(sex == "Female") %>% 
  ggplot(aes(height)) + geom_density(fill="pink", adjust = 2) #adjust the smoothness
```

```{r}
heights %>% filter(sex == "Male") %>%
  ggplot(aes(sample=height)) + geom_qq()
```

```{r}
heights %>% filter(sex == "Male") %>% ggplot(aes(sample=scale(height))) +
  geom_qq()+
  geom_abline()
```

### Data visualisation in practice

```{r}
data(gapminder)
gapminder %>% as_tibble()
```
Comparison of child mortality in five countries

1. Sri Lanka versus Turkiye
2. Poland versus South Korea
3. Pakistan versus Vietnam
4. Malaysia versus Russia
5. Thailand versus South Africa

```{r}
gapminder %>% filter(year == 2015 & country %in% c("Sri Lanka", "Turkey", 
                                                   "Poland", "South Korea", 
                                                   "Pakistan", "Vietnam",
                                                   "Malaysia", "Russia", 
                                                   "Thailand", "South Africa")) %>%
  select(country, infant_mortality)
```

```{r}
gapminder %>% filter(year == 2015 & country %in% c("Poland", "South Korea")) %>%
  select(country, infant_mortality)
```

```{r}
filter(gapminder, year==1962) %>% 
  ggplot(aes(fertility, life_expectancy, color=continent))+
  geom_point()
```
 
 _Faceting_
```{r}
filter(gapminder, year %in% c(1962, 2012)) %>% 
  ggplot(aes(fertility, life_expectancy, color=continent))+
  geom_point()+
  facet_grid(continent~year)
```
 
```{r}
filter(gapminder, year %in% c(1962, 2012)) %>% 
  ggplot(aes(fertility, life_expectancy, color=continent))+
  geom_point()+
  facet_grid(.~year)
```

`facet_wrap`

```{r}
years <- c(1962, 1980, 1990, 2000, 2012)
continents <- c("Europe", "Asia")
gapminder %>%
  filter(year %in% years & continent %in% continents) %>%
  ggplot(aes(fertility, life_expectancy, col=continent)) +
  geom_point()+
  facet_wrap(~year)
```

```{r}
gapminder %>% filter(year %in% c(1962, 2012))%>%
  ggplot(aes(fertility, life_expectancy, col=continent)) +
  geom_point()+
  facet_wrap(.~year, scales = "free")
```

_Time series plots_

```{r}
gapminder %>% 
  filter(country == "United States") %>%
  ggplot(aes(year, fertility))+
  geom_line(colour="red")
```

```{r}
countries <- c("South Korea", "Germany")

gapminder %>% filter(country %in% countries) %>% 
  ggplot(aes(year, fertility, col = country))+ 
  geom_line()
```

Labels instead of legends

```{r}
labels <- data.frame(country=countries, x = c(1975, 1965), y = c(60, 72))

gapminder %>%
  filter(country %in% countries) %>% 
  ggplot(aes(year, life_expectancy, col=country))+
  geom_line()+
  geom_text(data = labels, aes(x,y, label=country), size=3)+
  theme(legend.position = "none")
```

```{r}
gapminder <- gapminder %>% mutate(dollars_per_day = gdp/population/365)
past_year <- 1970
gapminder %>% filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day))+
  geom_histogram(binwidth = 1, fill="purple", color = "black")
```

```{r}
gapminder <- gapminder %>% mutate(dollars_per_day = gdp/population/365)
past_year <- 1970
gapminder %>% filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(log2(dollars_per_day)))+
  geom_histogram(binwidth = 1, fill="cyan", color = "black")
```

```{r}
filter(gapminder, year == past_year) %>%
  summarise(min = min(population), max = max(population))
```

```{r}
gapminder %>%
  filter(year == past_year) %>% 
  ggplot(aes(log10(population)))+
  geom_histogram(binwidth = 0.5, fill="red" ,color = "green")
```

```{r}
gapminder <- gapminder %>% mutate(dollars_per_day = gdp/population/365)
past_year <- 1970
gapminder %>% filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(dollars_per_day))+
  geom_histogram(binwidth = 1, color = "black") +
  scale_x_continuous(trans = "log2")
```

```{r}
gapminder <- gapminder %>% 
  mutate(group = case_when(
    region %in% c("Western Europe", "Northern Europe","Southern Europe",
                  "Northern America","Australia and New Zealand") ~ "West",
    region %in% c("Eastern Asia", "South-Eastern Asia") ~ "East Asia",
    region %in% c("Caribbean", "Central America",
                  "South America") ~ "Latin America",
    continent == "Africa" & region != "Northern Africa" ~ "Sub-Saharan",
    TRUE ~ "Others"))
```


```{r}
gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  mutate(region = reorder(region, dollars_per_day, FUN = median)) %>%
  ggplot(aes(dollars_per_day, region))+
  geom_point()+
  scale_x_continuous(trans="log2")
```

```{r}
p <- gapminder %>% 
  filter(year == past_year & !is.na(gdp)) %>%
  ggplot(aes(group, dollars_per_day)) +
  geom_boxplot()+
  scale_x_continuous(trans = "log2")+
  xlab("")+
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
p
```

